---
title: "1. t-tests"
author: "Anna Serrichio"
date: "2026-01-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction to t-tests

## What is a t-test?

A t-test compares the means of two groups to determine if they are statistically different from each other. In aphasia research, we use t-tests to answer questions like:

- Did naming accuracy improve from pre-treatment to post-treatment?
- Do people with Broca's aphasia have different working memory scores than people with Wernicke's aphasia?
- Is there a difference in treatment response between intensive vs. distributed therapy schedules?

## When to Use t-tests

Use t-tests when:

- You have **one continuous dependent variable** (e.g., naming accuracy %, reaction time, test scores)
- You want to compare **two groups or two time points**
- Your data is **approximately normally distributed** (or you have n > 30 per group)
- You have **independent observations** (each participant contributes one data point per condition)

## Types of t-tests

### 1. Independent Samples t-test (Between-subjects)
Compares two different groups of participants.

**Example:** Comparing naming scores between people with fluent vs. non-fluent aphasia.

### 2. Paired Samples t-test (Within-subjects)
Compares two measurements from the same participants.

**Example:** Comparing pre-treatment vs. post-treatment naming accuracy in the same individuals.

### 3. One-Sample t-test
Compares a sample mean to a known population value.

**Example:** Comparing your aphasia sample's cognitive scores to published norms (mean = 100).

---

# Required Packages

```{r packages}
# Install packages if needed (uncomment the lines below)
# install.packages("tidyverse")
# install.packages("effectsize")
# install.packages("rstatix")
# install.packages("ggplot2")
# install.packages("ggpubr")

# Load packages
library(tidyverse)    # Data manipulation and visualization
library(effectsize)   # Calculate effect sizes (Cohen's d)
library(rstatix)      # Pipe-friendly stats functions
library(ggplot2)      # Advanced plotting
library(ggpubr)       # Publication-ready plots
```

---

# Data Structure Requirements

## For Independent Samples t-test

Your data should be in **long format** with:
- One row per participant
- One column for the grouping variable (factor)
- One column for the outcome variable (numeric)

```{r indep-data-structure}
# Example structure for independent samples
independent_data <- data.frame(
  participant_id = paste0("P", 1:20),
  aphasia_type = rep(c("Fluent", "Non-fluent"), each = 10),
  naming_accuracy = c(rnorm(10, mean = 65, sd = 12),  # Fluent group
                      rnorm(10, mean = 52, sd = 15))  # Non-fluent group
)

head(independent_data)
```

## For Paired Samples t-test

Your data can be in **wide format** (one row per participant, separate columns for each time point) or **long format** (one row per observation).

```{r paired-data-structure}
# Wide format (easier for paired t-test)
paired_data_wide <- data.frame(
  participant_id = paste0("P", 1:15),
  pre_treatment = rnorm(15, mean = 55, sd = 10),
  post_treatment = rnorm(15, mean = 68, sd = 12)
)

head(paired_data_wide)

# Long format (better for plotting)
paired_data_long <- paired_data_wide %>%
  pivot_longer(cols = c(pre_treatment, post_treatment),
               names_to = "timepoint",
               values_to = "naming_accuracy")

head(paired_data_long)
```

---

# Running t-tests

## 1. Independent Samples t-test

### Check Assumptions

```{r indep-assumptions}
# 1. Check normality with Shapiro-Wilk test (p > .05 = normal)
independent_data %>%
  group_by(aphasia_type) %>%
  shapiro_test(naming_accuracy)

# 2. Visual check with Q-Q plot
ggqqplot(independent_data, x = "naming_accuracy", facet.by = "aphasia_type")

# 3. Check homogeneity of variance (Levene's test: p > .05 = equal variances)
independent_data %>%
  levene_test(naming_accuracy ~ aphasia_type)
```

### Run the t-test

```{r indep-ttest}
# Basic t-test
t_result_indep <- independent_data %>%
  t_test(naming_accuracy ~ aphasia_type, var.equal = TRUE) %>%
  add_significance()  # Adds significance stars
```

### Calculate Effect Size

```{r indep-effect}
# Cohen's d (small = 0.2, medium = 0.5, large = 0.8)
cohens_d(naming_accuracy ~ aphasia_type, 
         data = independent_data)
```

### Visualize Results

```{r indep-plot, fig.width=7, fig.height=5}
# Create publication-ready plot
ggplot(independent_data, aes(x = aphasia_type, y = naming_accuracy, fill = aphasia_type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 2) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "white") +
  labs(title = "Naming Accuracy by Aphasia Type",
       x = "Aphasia Type",
       y = "Naming Accuracy (%)") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## 2. Paired Samples t-test

### Check Assumptions

```{r paired-assumptions}
# 1. Check normality of DIFFERENCES
paired_data_wide <- paired_data_wide %>%
  mutate(difference = post_treatment - pre_treatment)

shapiro_test(paired_data_wide$difference)

# 2. Visual check
ggqqplot(paired_data_wide, x = "difference")
```

### Run the t-test

```{r paired-ttest}
# Basic paired t-test
t_result_paired <- paired_data_long %>% 
  t_test(naming_accuracy ~ timepoint, paired = TRUE) %>%
  add_significance()
```

### Calculate Effect Size

```{r paired-effect}
# Cohen's d for paired samples
cohens_d(
  data = paired_data_wide,
  formula = post_treatment ~ pre_treatment,
  paired = TRUE
)
```

### Visualize Results

```{r paired-plot, fig.width=7, fig.height=5}
# Box plot with individual trajectories
ggpaired(paired_data_long, x = "timepoint", y = "naming_accuracy",
         color = "timepoint", palette = "jco",
         line.color = "gray", line.size = 0.4,
         point.size = 2) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4, 
               fill = "white", color = "black") +
  labs(title = "Treatment Effect on Naming Accuracy",
       subtitle = "Lines connect the same participant across timepoints",
       x = "Timepoint",
       y = "Naming Accuracy (%)") +
  theme_minimal()
```

---

## 3. One-Sample t-test

### When to Use

Use when comparing your sample to a known population value, such as:
- Published normative data
- Chance performance level
- Theoretical baseline

```{r one-sample}
# Example: Are our participants' working memory scores different from 
# the normative mean of 100?

wm_scores <- data.frame(
  participant_id = paste0("P", 1:20),
  wm_score = rnorm(20, mean = 92, sd = 15)
)

# Test against population mean of 100
t_result_one <- t.test(wm_scores, mu = 100)
print(t_result_one)

# Effect size
cohens_d(
  data = wm_scores,
  formula = wm_score ~ 1,
  mu = 100
)
```

---

# Reporting Results

## APA Style Reporting Template

For an **independent samples t-test**:

> An independent samples t-test revealed that participants with non-fluent aphasia (M = XXX, SD = XXX) had significantly lower naming accuracy than participants with fluent aphasia (M = XXX, SD = XXX), t = [insert t value], p = [insert p value], d = [insert Cohen's d].

For a **paired samples t-test**:

> A paired samples t-test indicated that naming accuracy significantly improved from pre-treatment (M = XXX, SD = XXX) to post-treatment (M = XXX, SD = XXX), t = [insert t value], p = [insert p value] d = [insert Cohen's d].

---

# Common Issues and Solutions

## Issue 1: Non-Normal Data

**Solution:** Use the non-parametric **Wilcoxon test** instead (covered in the next tutorial).

```{r wilcoxon-preview}
# For independent samples: Mann-Whitney U test
wilcox.test(naming_accuracy ~ aphasia_type, data = independent_data)

# For paired samples: Wilcoxon signed-rank test
wilcox.test(paired_data_wide$post_treatment,
            paired_data_wide$pre_treatment,
            paired = TRUE)
```

## Issue 2: Unequal Variances

**Solution:** Use Welch's t-test (doesn't assume equal variances).

```{r welch}
# Set var.equal = FALSE
t.test(naming_accuracy ~ aphasia_type, 
       data = independent_data,
       var.equal = FALSE)  
```

